Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-05-03 doc/paper.Rmd
fix citations

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others? -->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

<!--# CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html or this: https://mapbiomas.org/en -->

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
For instance, <!--# parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50--100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable---particularly for plants, which are unable to move in response to changes in local conditions.
In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

<!--# Add here what is known about frag and drought effects - without setting it up so people expect a pval for an interaction term -->

Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\

@@

#### Vocab

This seems tacked on here and perhaps better in the discussion: populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].

\< !--
\# Gaoue = ?
Orou G. Gaoue, Carol C. Horvitz, Ulrich K. Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
-- \>

Delayed effects, time lag, the lag...avoid lagged effects

#### Citations / information to allocate to appropriate paragraphs

-   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesovaBelowgroundPlantFunctional2018.
    Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiroEffectRhizomeExposure2021

-   the observed effects on reproduction can have effects beyond demography, eg.
    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)

-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975

-   @rundelStructuralPhysiologicalAdaptation1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
    Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
    See also @rundelFunctionalTraitsBroadleaved2020\

    @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet

-   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
    But *Heliconia* times reproduction so seeds only germinate at start of rainy season.

-   refs in @leitoldNinoDroughtIncreased2018 peak in tree mortality can happen up to 3 years after drought event in experimental studies.
    Observational studies show mortality happening in same year as 2007 and 2010 droughts.

-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
    Similar increase in continuous and (large) fragments.
    Increase in turnover more variable in fragments.
    Coarse woody debris increased 62% during drought.\

<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).
These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest.
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992].
For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought.
Despite this, models of population dynamics have rarely included the effects of environmental drivers.
This has largely been due to the challenge---both ecologically and statistically---of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021].
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, [delete this phrase?
the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 <!--# weeks? --> prior to the appearance of infloresences.
Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering.
While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.

Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
For example, the mortality rate of tropical trees can spike in the year following a severe drought.
The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.

Finally, demographic delays could be an artefact of the timing of responses vis-à-vis the census date.
If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g. in Figure \@ref(fig:surv-spei)b).
In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
[consequence of demography on an annual scale and weather on a monthly / seasonal scale.
This problem not unique to us, same for everyone.
Mismatched scale at which analyses operate]

# Implications of delayed effects of climate on demography

Both extreme values and variation in SPEI are predicted to increase as climate change progresses ( <!--# cite IPCC? -->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.

@@

Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H. acuminata* growth and survival in fragments compared to continuous forest.
It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1-ha fragments.
Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.

-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
    Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon.
    Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
    Dry season droughts result in green-up, but not wet season droughts.
    During drought, old leaves are shed and new leaves are maintained.
    2-5 months for newly flushed leaves to fully expand.

-   Ecological Consequences of these patterns: an eye towards a drier/wetter future

@@

-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
    Smaller/larger plants might be more susceptible to drought.
    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
    The `dlnm` package does not currently allow for modeling of interactions, and we lacked data on plant size at the same time scale as SPEI (monthly), so interactions with size were not included in this analysis.

-   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance. While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood. They could act in concert to exacerbate or dampen population dynamics."

-   few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
    This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
    <!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->

-   More experiments like westerband that try and see how contiions today influence growth in future years Generalizations about the direction and magnitude of delayed demographic responses will continue to prove elusive.
    Without experiments designed greater understanding of the complex ways in which light, environmental conditions, and other ecological conditions interactions to influence plant vital rates (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002),

-   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.

@@

>>>>

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015a].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @bruna2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@laurance2001; @opdam2004, @selwood2015].

Whether plant populations in tropical forest fragments are more susceptible to drought remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009] <!--# Emilio had written "Gaoue" as potential reference, but the on I have in Zotero is about trees, not understory plants -->.
We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\

@@

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).

These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest.
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006]. There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].

Despite this, models of population dynamics have rarely included the effects of environmental drivers.
This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

However, while it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@evers2021], and hence be shaped by climatic events at any point in that physiological window.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering in one year is associated with increased reproduction and growth in the next. **(is this an acceptable colloquial way to interpret the covariate effect?)**.

Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [sensu @pumisutapon2012].
The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimeová2018].
This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@bruna2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
This would also be consistent with the results of prior experiments, in which the growth rates of *H. acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@bruna2005].

Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021].
For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994].
This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.

Finally, demographic delays could be an artefact of the timing of responses vis-à-vis the census date.
If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g. in Figure \@ref(fig:surv-spei)b).
In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
This is possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
While the very slow growth and extremely low mortality rates of *H. acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated.

***THIS INFO PROB NEEDS TO BE WORKED IN SOMEWHERE ABOVE***

@@

@@

Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@laurance2014].
However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @bruna2005a; @markewitz2010; @westerband2015; @westerbandEarlyLifeConditions2017)

Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms (cite `dlnm`).
They also don't allow for modeling interactions between \_\_\_\_ and plant size; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019], the ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.

Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @evers2021].
Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@bruna2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

@@

### Left-over text

-   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
-   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is. It We hypothesize that the microclimate in continuous forest plots is more stable than that in 1 ha fragments, and that this stability will demographically buffers populations in forest interiors [@brunaExperimentalAssessmentHeliconia2002].
-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
-   eg. gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets. Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon. Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought. Dry season droughts result in green-up, but not wet season droughts. During drought, old leaves are shed and new leaves are maintained. 2-5 months for newly flushed leaves to fully expand

#### Citations / information could allocate to appropriate paragraphs

-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet

-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
    Similar increase in continuous and (large) fragments.
    Increase in turnover more variable in fragments.
    Coarse woody debris increased 62% during drought.
